---
title: An introduction to R for phylogenetics
author: François Michonneau
date: December 7th, 2015
output: pdf_document
---

# Introduction

## Why R?

Are you working on more than 100 species? Are you going to analyze more than 3
markers? Are you going to report bootstrap values on a tree with hundreds of
nodes? If you answer "yes" to any of these questions, you're going to need the
help of a computer to keep your sanity.

Learning a programming language is not easy. However, the benefits of your time
investment will be higher than you can imagine. For instance, it will give you
new perspectives on the kind of analyses you'll be able to perform. You'll also
be able to tailor the methods you want to use to your specific questions instead
of being constrained by the way the author of a software package designed them.

R is a really popular programming language because it can input and output many
type of data, it's really robust, it has great analytical and graphical
capabilities built-in, and there is a huge community of users willing to help.

The goal of this short handout is to show you some basic examples of data
manipulation. I hope these short general examples will spark your interest to
learn more about R as a programming language. I prefer not to cover specific
methods that might be applicable to just a small group of people, but instead
give a quick overview of basic functionalities that can be applied to data
management and manipulation, something everybody struggles with.

## What is R?

R is both a programming language and an environment for statistical and
graphical analysis. It is open-source (you can use it for free, you can look at
the source code, you can modify it, you can redistribute it) and cross-platform
(you can use it on any computer).

R is very flexible and its functionalities can be extended with its package
system. R is used to analyze data of any kind (biological, geographical,
medical, financial) and of almost any size.

Users interact with R by writing commands. Each command uses one or more
functions, and a set of commands forms a script. It's also very easy to write
your own functions, and you may incorporate them within your scripts.


## Advantages of using R

### Reproducibility

One of the main advantages of R is that **reproducibility** is built-in. The
system of scripts makes it easy to reproduce all the steps used in your
analysis. In a point-and-click application, it might be hard to remember which
options you used in your analyses. With R, you can keep track of all the steps
you used. R also allows you to automate many repetitive tasks. For instance, if
you get new samples and you want to re-run an analysis to include them, with R
you will just need to update your original dataset and re-run the scripts you
wrote when your dataset was smaller.

### Sharing

The reproducibility also allows you to share your scripts easily. If you provide
the data you used and your script to someone else, then your collaborator will
be able to obtain the exact same results. This is even made easier in R as it's
cross-platform (you don't have to worry if your collaborator will be using a
Mac, a PC or a linux computer) and also it does not require your collaborator to
buy an expansive piece of software.

### Integration

Because R is versatile, it can also be used at each step of the work flow. R can
import and export data in many different formats. Developing methods under a
single user interface greatly facilitates their integration and
inter-operability. Furthermore, because R can interact with many programming
languages, developers can use the language that is best suited for the software
they develop, while users do not have to learn a new interface.

R has a steep learning curve but investing time and energy to learn it is very
worthwhile. R allows you to manage, rename, subset, bind data sets in a few
commands. As data sets become larger, doing this kind of operations by hand
would take hours and it would be easy to introduce mistakes that would not
necessarily be noticed.



## Getting started

### Where do I get R?

R can be downloaded from the R-project website: <http://www.r-project.org/>.

I also encourage you to install RStudio <https://rstudio.com/>.

### Where do I find help?

As R is becoming prevalent there are a number of resources that have been
developed to help new users getting started.

For general help:

* **[R for cats](http://rforcats.net/)** an excellent overview to get started.
* **[An introduction to R](http://cran.r-project.org/doc/manuals/R-intro.pdf)**
  A comprehensive and easy-to-follow tutorial produced by the R Development Core
  Team (i.e., the people who are writing R). There are other manuals available,
  some of them are not for beginners but are full of useful information once you
  know your way around R.
* **[R for Beginners](http://cran.r-project.org/doc/contrib/Paradis-rdebuts_en.pdf)**
   A tutorial by Emmanuel Paradis.
* **[The R FAQ](http://cran.r-project.org/doc/FAQ/R-FAQ.html)**

Courses:

* **[R programming for biologists](http://r-bio.github.io)**
* **[R lessons](http://www.datacarpentry.org/R-ecology/)** from Data Carpentry



For help about phylogenetics and comparative methods in R:

* Phylogenetic Comparative Methods Wiki Tutorials and overview of methods
  available in R for phylogenetic comparative analysis. The wiki grew out of a
  Hackathon on Comparative Methods in R held at the National Evolutionary
  Synthesis Center (NESCent) 10-14 December 2007.  <http://www.r-phylo.org/>
* Analysis of Phylogenetics and Evolution with R A book written by Paradis.
  This book is a very useful reference on how to do evolutionary analyses using
  the `ape` package, written by its maintainer. It is available from Springer
  and Amazon.
* The Bodega workshop on phylogenetics and comparative methods in R.
  <http://bodegaphylo.wikispot.org/Phylogenetics_and_Comparative_Methods_in_R>
* The R-sig-phylo mailing list:
  <https://stat.ethz.ch/mailman/listinfo/r-sig-phylo>. Even Felsenstein is on it
  and he replies to any questions about comparative methods in less than 5
  minutes usually!


### R basics

Being organized when using R is particularly important. When you import local
files, R looks for them in what is called the *working directory*. The best
way to work with R is thus to have a different directory for each analysis which
contains all the files that you will need.

In R, the command line starts with the symbol `>`. After typing a command, you
can execute it by pressing `Enter`. If the command is incomplete, the command
entry is continued on the following line which then starts with the symbol
`+`. R is nonsensitive to space characters (e.g., spaces, tab, new lines) but
it's case sensitive (e.g, `Hello` and `hello` are different variable names). To
assign a value to a variable, you need to use the symbols `<-` (`Alt+-` in
RStudio). In R, the comment character is `#`, R ignore anything found after it.

Let's take a look at a small example to understand how to use functions and
variables.

```{r, keep.source=TRUE}
## This creates a variable (x) which contains number 0 to 10
x <- 0:10
## Just typing the name of the variable gives you its content
x
## Let's calculate the mean value of the numbers from 0 to 10
mean(x)
## It's also possible to store the value of the mean in a
## new variable
y <- mean(x)
## You can do operations on variables, for instance
y + 1
y * 3
z <- y^2
z
```

With R it's also possible, in some cases, to directly create or change
attributes of a variable. Here is an example where we use species names as
attributes for data about the length of a group of species.

```{r, keep.source=TRUE}
## Let's create a vector of 5 elements: the species names
## of our group
spNames <- c("edulis", "impatiens", "atra", "coluber", "fuscopunctata")
spNames
## Let's now assume that we have measurements for
## each of these species (in cm)
spLength <- c(12, 8, 23, 37, 42)
spLength
## Now we are going to use the species names as an attribute
## of the length of these species
names(spLength) <- spNames
## Let's have a look at the result
spLength
```

Note that we used `"` on each side of our species names. Otherwise, R would
have interpreted them as variable names (`y` was a variable name in the
previous example), and because they don't exist (we didn't define them) we would
have gotten an error message.

Now, let's say that we would like to add the first letter of the genus name to
these species names. We could do this in two ways. We could first recreate the
`spNames` vector with the new names, but R allows you to do this much more
efficiently:

```{r, keep.source=TRUE}
newSpNames <- paste("H.", spNames, sep="")
newSpNames
names(spLength) <- newSpNames
spLength
```

With only 5 species you could have done it pretty quickly by hand, but imagine
all the time this would have saved you if you had 1,000 species to rename.

Now, imagine that in addition of length data, we also have the depth (in meters)
at which all these species live and their colors. We can store all
these data in R in a particular class of object called a data frame.

```{r, keep.source=TRUE}
dataSp <- data.frame(spLength,
                     spColors=c("pink", "purple", "black", "black", "brown"),
                     spDepth=c(12, 2, 1, 1, 15))
rownames(dataSp) <- newSpNames
dataSp
```

An important aspect of data manipulation is sub-setting. Using the square
brackets allows you to select one or more elements of a vector. Here are some
examples:

```{r, keep.source=TRUE}
## Sub-setting one element of a vector using a numeric index:
spLength[2]
## Sub-setting one element of a vector using its name:
spLength["H.edulis"]
## Sub-setting several elements of a vector:
spLength[c(1,3)]
spLength[2:4]
## Sub setting from a data frame:
## the first element is the row index,
## the second the column index
dataSp[1,3]
dataSp["H.atra", "spDepth"]
dataSp["H.impatiens", ]
```

In the latter case we left the second dimension empty, so all columns are
returned.

# Importing and exporting data

The main purpose of R is to import your data to manipulate and analyze it. In
this section we will (1) import data, (2) manipulate them to create a new
dataset, (3) export this new dataset.

We'll use two datasets. One has the size of three color morphs in a species
complex (`lengthData`), the other has the depth at which the individuals
were found (`depthData`). The individuals in each case are uniquely
identified by a number, their `fieldID`. We are going to merge these two
data frames together and export the result into a new file.

```{r, keep.source=TRUE}
## Load into memory two datasets, stored as CSV files
lengthData <- read.csv(file="http://goo.gl/pHweF")
depthData <- read.csv(file="http://goo.gl/Vc4Ov")
## We can inspect the beginning of these two files with the
## head() function
head(lengthData)
head(depthData)
## Let's merge them
combinedData <- merge(lengthData, depthData, by="fieldID", all=TRUE)
## Let's look at the result
head(combinedData)
## We can now export it, this will create the 'combinedData.csv'
## in your working directory
write.csv(combinedData, file="combinedData.csv")
```



# The R packaging system

You can extend the "basic" functionalities offered natively by R, by installing
packages. As of today, there are 7581 packages available. You can install
packages from the R GUI, or if you know the name of the package you want to
install (in this example the package `ape`), you can type at the command line:

```{r, eval=FALSE}
install.packages("ape")
```

By default, you will download the packages from one of the CRAN (Comprehensive R
Archive Network) mirrors. Mirrors are identical copies of a given website. They
reduce the load on individual servers, and users can enjoy faster connections by
using a server located in their geographical area. When asked to choose a CRAN
mirror, just pick one located in the US.

```{r, eval=FALSE}
install.packages("phylobase")
```

# R packages for phylogenetics

The number of R packages that have been developed for phylogenetics has been
increasing rapidly recently. Below is the list of some of the main packages. You
can find a more comprehensive list on the
[CRAN Task View page](http://cran.r-project.org/web/views/Phylogenetics.html).


* **`ape`**: (Analysis of Phylogenetics and Evolution) is maintained by Emmanuel
  Paradis. It's the most comprehensive package for phylogenetics. It provides
  functions to import data (sequences and trees) in various formats (e.g.,
  phylip, nexus, newick, genbank). There are a few functions to infer
  phylogenies within R (neighbor-joining and maximum likelihood with
  phyML). `ape` can be used for tree manipulation (pruning and grafting) and
  display. `ape` implements various methods for comparative and diversification
  analyses as well as ancestral state reconstruction.

* **`phylobase`**: This package provides a common suite of basic functions for
  manipulating and storing trees and data. As `ape`, it provides functions for
  importing tree (from NEXUS and Newick files) and pruning trees. The objects
  are stored in a S4 class which provides validation on the objects, so the user
  can't mess them up. The functions for tree manipulation are faster than in
  `ape`. The main advantage of `phylobase` is that it allows the user to
  manipulate easily data associated with tips or nodes of a tree, which
  facilitates the implementation of comparative methods.

* **`seqinr`**: Seqinr provides a lot of functions to manipulate and analyze
  sequence data. It can import sequences from a variety of formats, converts
  nucleotides to amino-acids, etc.

* **`picante`** picante provides various tools and methods for community
phylogenetics, including an interface with phylocom.

* **`geiger`** Geiger provides functions to use comparative methods in the
context of adaptive radiations

* **`bamm`** can be used for diversification analyses.
